Supp mat: time series paper

### 2017-07-07 writing/manuscript_draft.tex
added section to reference figure in supp mat

Observational studies and population censuses have become a cornerstone
of modern ecological research and conservation biology (Magurran et al.
2010). Longterm data is necessary to both understand population dynamics
and to assess extinction risk of species. Even though many time series
may now be considered ``longterm'', most are still short (i.e.~less than
X years). \emph{I could have statistic on how long is considered
``longterm'' in papers}. Time series are typically short for a variety
of reasons. They are often coupled with an experiment, which may only
last a couple of years. In addition, short funding cycles make it
difficult to examine populations over longer periods of time
(citations). There are, of course, notable exceptions including the Isle
Royale study, Iowa corn study, cod, lynx-hair, and many others
(citations).

@@

In determining the number of samples required for a particular
experiment four quantities are intricately linked: statistical power
(1-\(\beta\)), effect size, \(\alpha\), and sample size (Legg and Nagy
2006). The exact relationship between these quantities depends on the
specific statistical test. Formally, statistical power (\(1-\beta\)) is
one minus the probability of a type II error (\(\beta\)), or false
negative. In other words, it is a measure of how well you could detect a
trend given it is actually significant. Prior to an experiment, one
could set appropriate levels of power, significance threshold
(\(\alpha\)), and the effect size to estimate the sample size required
for the experiment. This approach, however, is not straight-forward for
a time series, or more complicated scenarios (P. C. D. Johnson et al.
2015), as data points are clearly non-independent.

For time series data, two general approaches can be used. First, as can
be done with experimental data, we can relax the definition of
statistical power (cite Bolker). Instead of using classic statistical
tests (e.g.~t-test), and corresponding formulas for determinig sample
size, simulation approaches can be used (Gerrodette 1987). Simple
population models can be simulated with parameter values corresponding
to a population of interest. Out of a large number of simulations, power
can be calculated as the fraction of simulations that meet some
criteria. The specific criteria depend on the question at hand. Then
time series simulations of different lengths are run to determine the
minimum time series length required to be confident that you will be
able to answer your question (cite Bolker). In addition to using
simulated data, empirical time series can also be used. We usually do
not have multiple replicates of similar populations, but we can
subsample an empirical time series (cite). Subsamples of different
lengths can then be evaluated to see which fraction of subsamples meet
some criteria, again a measure of statistical power. Similar to the
simulation approach, we can use these measures of power to determine the
minimum time series required for a particular question of interest.

Past work has investigated questions related to the minimum time series
required to estimate trends in population size over time (Giron-Nava et
al. 2017). For example, Rhodes and Jonzen (2011) examined the optimal
allocation of effort between spatial and temporal replicates. Using
simple populations models, they found that the allocation of effort
depends on environmental variation, spatial and temporal
autocorrelation, and observer error. Rueda-Cediel et al. (2015) also
used a modeling approach, but parameterized it for a threatened snail,
\emph{Tasmaphena lamproides}. They found that for this short-lived
organims, that 15 years was adequate to assess longterm trends in
abundance. However, these studies, and other past work, have focused
either on theoretical aspects of monitoring design or focused on single
species.

In this paper, we use both simulations and a database of empirical time
series to determine the minimum number of years required to address
various questions. More specifically, we estimate the minimum time
series length required to asssess longterm changes in abundance via
simple linear regression.

@@

One approach to determining the minimum time series length needed is
through repetitive simulations of a population model (Gerrodette 1987).
This is the same approach one might use in sample size calculations for
any experimental design too complicated for simple power analyses (P. C.
D. Johnson et al. 2015) (and cite Bolker). We only briefly discuss this
approach as it has been desribed elsewhere (citations). Essentially, we
take a stochastic population model and simulate it for a number of years
repetitively. We can then determine the minimum time series required to
achieve designed levels of significance and power for a particular
question. This approach requires us to determine values for our model
parameters (e.g.~growth rate). As an example, we can take the following
population model for population size \(N\) at time \(t\): \[
N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\mu, \sigma)
\] where \(\epsilon\) is a normally-distributed random noise term with
mean \(\mu\) and standard deviation \(\sigma\). The rate of growth \(r\)
is also the trend strength of increase or decrease.

By simulating this model repetitively for different lengths of time, we
can determine the minimum time series required. Here statistical power
is simply the fraction of simulations with significant trends in
abundance from simple linear regression. Then, the minimum time required
is the minimum time required to obtain statistical power above some
pre-determined threshold (historically at 0.8).

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{(a) example of simulated time series for 40 time periods and
(b) statistical power versus the simulated time series length. The
horizontal, dashed line is our desired statistical power of 0.8. The
vertical, dashed line is the minimum time required to achieve our
desired statistical power. (c) Minimum time required for different
values of the trend strength (\(r\)) and (d) minimum time required for
different levels of population variability
(\(\sigma\)).\label{fig:theoretical_approach}}
\end{figure}

@@

We use a database of 2444 population time series compiled in (Keith et
al. 2015). The data is originally from the Global Population Dynamics
Database (NERC Centre for Population Biology 2010). We filtered out
short time series (less than 35 years), and those with missing data, and
were left with 878 time series. The database includes information on 478
vertebrate species with a focus on mammals, birds, and fish. The
database also includes information on generation length and census
specifications. For each time series, we also calculated variables of
interest like variance in population size, longterm trend in abundance
(slope coefficient from simple linear regression), and temporal
autocorrelation. Lastly, we also calculated the minimum number of years
required (10 years or 3 generations) under IUCN Red List Criterian A2 to
classify a species as vulnerable (IUCN 2012).

For a subset of populations in the database (\(n =\) 547), we were able
to connect a database on biological characteristics (\textbf{cite
database here}) like body size and generation length. All 547
populations are within the Aves class. In the supplementary material, we
explore how the minimum time required does not strongly correlate with
any biological characteristics.

@@

We assume that each time series in our database is long enough to
include all necessary information (e.g.~variability) about the
population. In other words, each time series is a representative sample.
We first take all possible contiguous subsamples of each time series.
For example, a time series of 40 years would have 39 possible contiguous
subsamples of length 2, 38 possible contiguous subsamples of length 3,
and continuing until 1 possible contiguous subsample of length 40
(Gerber, DeMaster, and Kareiva 1999, Giron-Nava et al. (2017)). Then we
determined the \(T_{min}\) required to be confident (i.e.~have high
enough statistical power) in an estimate of the slope coefficient from
linear regression. Therefore, we run a linear regression for each
subsampled time series. Then, we determine the fraction of subsamples of
a particular length that have estimated slope coefficients which are
statistically different from zero given the longterm, or ``true``, time
series also has a significant slope. This is a measure of statistical
power (often denoted as \(1- \beta\), where \(\beta\) is the probability
of a type II error). Lastly, we determine which subsample length is
required to guarantee we have above a certain threshold of statistical
power, which we will assume is 0.8 by convention (citation). A power of
0.8 would imply that given a trend exists, there is a 0.8 probability of
detecting the trend. Complementary, there would be a 0.2 probability of
a false negative.

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{(a) Population size of Harp Seals over time. The line is the
best fit line from linear regresssion. (b) Statistical power for
different subsets of the time series in panel
a.\label{fig:empirical_approach_example}}
\end{figure}

We also examine how minimum time series calculations changed depending
on the question of interest. In the supplementary material, we show how
to make similar calculations to determine the minimum time series
required to estimate overall population growth rate, the variability in
population size, and future extinction risk. Again, we took subsamples
of time series data to determine \(T_{min}\) for each of these
questions.

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Distribution of the minimum time required in order to detect a
significant trend (at the 0.05 level) in abundance given power of
0.8.\label{fig:min_time_dist}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-6-1.pdf}
\caption{Minimum time required to estimate change in abundance
correlated with (a) trend strength (absolute value of slope coefficient
estimated from linear regression), (b) population variance (interannual
variability in population size), and (c) temporal
autocorrelation.\label{fig:correlates}}
\end{figure}

@@

We explored two different approaches to estimate the minimum time series
length required to address a particular question of interest. We
specifically asked, what is the minimum time series length required to
determine longterm changes in abundance using linear regression? The
simulation-based approach has been suggested by others, especially in
situations more complicated than that suited for classic power analysis
(Gerrodette 1987, Johnson2015). Our simulations support past work that
longer time series are needed when the trend strength (i.e.~rate of
increase or decrease) is weak or when population variability is high
(Gerrodette 1987).

@@

There has also been interest in determining the minimum time series
length in the bird community (Hatch 2003, ({\textbf{???}}), Seavy and
Reynolds (2007)). Hatch (2003) used seabird monitoring data to estimate
the minimum time needed to achieve appropriate levels of power. He found
that the minimum time required ranged between 11 and 69 years depending
on species, trend strength, and study design. Similarly,
({\textbf{???}}) recommended a standard of 80\% power to detect a 50\%
decline over 20 years in landbird populations. They found that only
about 42\% of populations in the Breeding Bird Survey met their
criteria. Seavy and Reynolds (2007) asked whether statistical power was
even a useful framework for assessing longterm population trends. They
used 24 years of census data on Red-tailed Tropicbirds
(\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50\%
decline over 10 years almost always resulting in high statistical power
(above 0.8). Therefore, they cautioned against only using power analyses
to design monitoring schemes and instead argued for precision based
metrics ..

An important related question, is the optimal allocation of sampling
effort in space versus time (citations). In a theoretical investigation
of this question, Rhodes and Jonzen (2011) found that the optimal
allocation of sampling depended strongly on temporal and spatial
autocorrelation. If spatial population dynamics were highly correlated,
then it was better to sample more temporally, and vice versa. Our work
supports this idea as populations with strong temporal autocorrelation
needed less years of sampling (cite figure here). Morrison and Hik
(2008) also studied the optimal allocation of sampling effort in space
versus time, but used emprical data from a longterm census of the
collared pika (\emph{Ochotona collaris}) found in the Yukon. They
estimated longterm growth rates among three subpopulations over a
10-year period. They found that censuses less than 5 years may be
misleading and that extrapolating from one population to another, even
when nearby, may be difficult. Therefore, they argued more sampling
should be conducted spatially than temporally (Morrison and Hik 2008).

@@

We use a database of 878 populations to determine the minimum time
series length required. We show that to identify longterm changes in
abundance, 10-15 years of continuos monitoring are required (Fig.
\ref{fig:min_time_dist}). In line with theoretical predictions, we also
show that \(T_{min}\) is strongly correlated with the overall trend in
abundance, variability in abundance, and the temporal autocorrelation
(Fig. \ref{fig:correlates}). Our work implies that for many populations,
census data less than 10-15 years are probably not reliable. This
stresses the importance of longterm monitoring programs. From both a
scientific and management perspective estimates of \(T_{min}\) are
important. If a time series is too short, we are likely to make
incorrect conclusions about longterm trends. In addition, a time series
that is too long may be a poor use of already limited funds (Gerber,
DeMaster, and Kareiva 1999). Future work can examine other species, with
a wider range of life history characteristics. In addition, similar
approaches can be used to determine the minimum time series length
required to address other questions.

@@

\hypertarget{ref-Johnson2015}{}
Johnson, Paul C D, Sarah J E Barry, Heather M. Ferguson, and Pie Müller.
2015. ``Power analysis for generalized linear mixed models in ecology
and evolution.'' \emph{Methods in Ecology and Evolution} 6 (2): 133--42.
doi:\href{https://doi.org/10.1111/2041-210X.12306}{10.1111/2041-210X.12306}.

@@

>>>>

Observational studies and population censuses have become a cornerstone
of modern ecological research and conservation biology (Magurran et al.
2010). long-term data is necessary to both understand population
dynamics and to assess extinction risk of species. Even though many time
series may now be considered ``long-term'', most are still short
(i.e.~less than X years). \emph{I could have statistic on how long is
considered ``long-term'' in papers}. Time series are typically short for
a variety of reasons. They are often coupled with an experiment, which
may only last a couple of years. In addition, short funding cycles make
it difficult to examine populations over longer periods of time
(citations). There are, of course, notable exceptions including the Isle
Royale study, Iowa corn study, cod, lynx-hair, and many others
(citations).

@@

In determining the number of samples required for a particular
experiment four quantities are intricately linked: statistical power
(1-\(\beta\)), effect size, \(\alpha\), and sample size (Legg and Nagy
2006). The exact relationship between these quantities depends on the
specific statistical test. Formally, statistical power (\(1-\beta\)) is
one minus the probability of a type II error (\(\beta\)), or false
negative. In other words, it is a measure of how well you could detect a
trend given it is actually significant. Prior to an experiment, one
could set appropriate levels of power, significance threshold
(\(\alpha\)), and the effect size to estimate the sample size required
for the experiment. This approach, however, is not straight-forward for
a time series, or more complicated scenarios (P. C. Johnson et al.
2015), as data points are clearly non-independent.

For time series data, two general approaches can be used. First, as can
be done with experimental data, we can relax the definition of
statistical power (cite Bolker). Instead of using classic statistical
tests (e.g.~t-test), and corresponding formulas for determinig sample
size, simulation approaches can be used (Gerrodette 1987). Simple models
can be simulated with parameter values corresponding to a population of
interest. Out of a large number of simulations, power can be calculated
as the fraction of simulations that meet some criteria. The specific
criteria depend on the question at hand. Then, time series simulations
of different lengths are run to determine the minimum time series length
required to be confident that you will be able to answer your question
(cite Bolker). In addition to using simulated data, empirical time
series can also be used. We usually do not have multiple replicates of
similar populations, but we can subsample an empirical time series
(Gerber, DeMaster, and Kareiva 1999). Subsamples of different lengths
can then be evaluated to see which fraction of subsamples meet some
criteria, again a measure of statistical power. Similar to the
simulation approach, we can use these measures of power to determine the
minimum time series required for a particular question of interest.

Past work has investigated questions related to the minimum time series
required to estimate trends in population size over time (Giron-Nava et
al. 2017). For example, Rhodes and Jonzen (2011) examined the optimal
allocation of effort between spatial and temporal replicates. Using
simple populations models, they found that the allocation of effort
depends on environmental variation, spatial and temporal
autocorrelation, and observer error. Rueda-Cediel et al. (2015) also
used a modeling approach, but parameterized amodel specifically for a
threatened snail, \emph{Tasmaphena lamproides}. They found that for this
short-lived organim, 15 years was adequate to assess long-term trends in
abundance. However, these studies, and other past work, have typically
focused either on theoretical aspects of monitoring design or focused on
single species.

In this paper, we use both simulations and a database of empirical time
series to determine the minimum number of years required to address
various questions. More specifically, we estimate the minimum time
series length required to asssess long-term changes in abundance via
simple linear regression. We examine 878 population time series to
estimate this minimum time series length required.

@@

One approach to determining the minimum time series length needed is
through repetitive simulations of a population model (Gerrodette 1987).
This is the same approach one might use in sample size calculations for
any experimental design too complicated for simple power analyses (P. C.
Johnson et al. 2015) (and cite Bolker). We only briefly discuss this
approach as it has been desribed elsewhere (citations). Essentially, we
take a stochastic population model and simulate it for a number of years
repetitively. We can then determine the minimum time series required to
achieve designed levels of significance and power for a particular
question. This approach requires us to determine values for our model
parameters (e.g.~growth rate). As an example, we can take the following
population model for population size \(N\) at time \(t\): \[
N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\mu, \sigma)
\] where \(\epsilon\) is a normally-distributed random noise term with
mean \(\mu\) and standard deviation \(\sigma\). The rate of growth \(r\)
is also the trend strength of increase or decrease.

By simulating this model repetitively for different lengths of time, we
can determine the minimum time series required. Here statistical power
is simply the fraction of simulations with significant trends in
abundance. A significant trend in abundance is determined by estimating
the significance of the slope parameter from simple linear regression.
Then, the minimum time required (\(T_{min}\)) is the number of years
required to obtain statistical power above some pre-determined threshold
(historically at 0.8). Statistical power of 0.8 would indicate that, if
there was a true trend in abundance, we would have a 0.8 probability of
detecting the trend.

@@

@@

We use a database of 2444 population time series compiled in (Keith et
al. 2015). The data is originally from the Global Population Dynamics
Database (NERC Centre for Population Biology 2010). We filtered out
short time series (less than 35 years), and those with missing data, and
were left with 878 time series. The database includes information on 478
vertebrate species with a focus on mammals, birds, and fish. The
database also includes information on generation length and census
specifications. For each time series, we also calculated variables of
interest like variance in population size, long-term trend in abundance
(slope coefficient from simple linear regression), and temporal
autocorrelation. Lastly, we also calculated the minimum number of years
required (10 years or 3 generations) under IUCN Red List Criterian A2 to
classify a species as vulnerable (IUCN 2012).

For a subset of populations in the database (\(n =\) 547), we were able
to connect a database on biological characteristics (Myhrvold et al.
2015) like body size and generation length. All 547 populations are
within the Aves class. In the supplementary material, we explore how the
minimum time required does not strongly correlate with any biological
characteristics.

@@

We assume that each time series in our database is long enough to
include all necessary information (e.g.~variability) about the
population. In other words, each time series is a representative sample.
We first take all possible contiguous subsamples of each time series.
For example, a time series of 40 years would have 39 possible contiguous
subsamples of length 2, 38 possible contiguous subsamples of length 3,
and continuing until 1 possible contiguous subsample of length 40
(Gerber, DeMaster, and Kareiva 1999, Giron-Nava et al. (2017)). Then we
determined the \(T_{min}\) required to be confident (i.e.~have high
enough statistical power) in an estimate of the slope coefficient from
linear regression. Therefore, we run a linear regression for each
subsampled time series. Then, we determine the fraction of subsamples of
a particular length that have estimated slope coefficients which are
statistically different from zero given the long-term, or ``true``, time
series also has a significant slope. This is a measure of statistical
power (often denoted as \(1- \beta\), where \(\beta\) is the probability
of a type II error). Lastly, we determine which subsample length is
required to guarantee we have above a certain threshold of statistical
power, which we will assume is 0.8 by convention (citation).

We also examine how minimum time series calculations changed depending
on the question of interest. In the supplementary material, we show how
to make similar calculations to determine the minimum time series
required to estimate overall population growth rate (Fig.
\ref{fig:growth_rate}), the variability in population size, and future
extinction risk. Again, we took subsamples of time series data to
determine \(T_{min}\) for each of these questions.

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Distribution of the minimum time required in order to detect a
significant trend (at the 0.05 level) in abundance given power of
0.8.\label{fig:min_time_dist}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Minimum time required to estimate change in abundance
correlated with (a) trend strength (absolute value of slope coefficient
estimated from linear regression), (b) population variance (interannual
variability in population size), and (c) temporal
autocorrelation.\label{fig:correlates}}
\end{figure}

@@

We explored two different approaches to estimate the minimum time series
length required to address a particular question of interest. We
specifically asked, what is the minimum time series length required to
determine long-term changes in abundance using linear regression? The
simulation-based approach has been suggested by others, especially in
situations more complicated than that suited for classic power analysis
(Gerrodette 1987, Johnson2015). Our simulations support past work that
longer time series are needed when the trend strength (i.e.~rate of
increase or decrease) is weak or when population variability is high
(Gerrodette 1987).

@@

There has also been interest in determining the minimum time series
length in the bird community (Hatch 2003, ({\textbf{???}}), Seavy and
Reynolds (2007)). Hatch (2003) used seabird monitoring data to estimate
the minimum time needed to achieve appropriate levels of power. He found
that the minimum time required ranged between 11 and 69 years depending
on species, trend strength, and study design. Similarly,
({\textbf{???}}) recommended a standard of 80\% power to detect a 50\%
decline over 20 years in landbird populations. They found that only
about 42\% of populations in the Breeding Bird Survey met their
criteria. Seavy and Reynolds (2007) asked whether statistical power was
even a useful framework for assessing long-term population trends. They
used 24 years of census data on Red-tailed Tropicbirds
(\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50\%
decline over 10 years almost always resulting in high statistical power
(above 0.8). Therefore, they cautioned against only using power analyses
to design monitoring schemes and instead argued for precision based
metrics ..

An important related question, is the optimal allocation of sampling
effort in space versus time (citations). In a theoretical investigation
of this question, Rhodes and Jonzen (2011) found that the optimal
allocation of sampling depended strongly on temporal and spatial
autocorrelation. If spatial population dynamics were highly correlated,
then it was better to sample more temporally, and vice versa. Our work
supports this idea as populations with strong temporal autocorrelation
needed less years of sampling (cite figure here). Morrison and Hik
(2008) also studied the optimal allocation of sampling effort in space
versus time, but used emprical data from a long-term census of the
collared pika (\emph{Ochotona collaris}) found in the Yukon. They
estimated long-term growth rates among three subpopulations over a
10-year period. They found that censuses less than 5 years may be
misleading and that extrapolating from one population to another, even
when nearby, may be difficult. Therefore, they argued more sampling
should be conducted spatially than temporally (Morrison and Hik 2008).

@@

We use a database of 878 populations to determine the minimum time
series length required. We show that to identify long-term changes in
abundance, 10-15 years of continuos monitoring are required (Fig.
\ref{fig:min_time_dist}). In line with theoretical predictions, we also
show that \(T_{min}\) is strongly correlated with the overall trend in
abundance, variability in abundance, and the temporal autocorrelation
(Fig. \ref{fig:correlates}). Our work implies that for many populations,
census data less than 10-15 years are probably not reliable. This
stresses the importance of long-term monitoring programs. From both a
scientific and management perspective estimates of \(T_{min}\) are
important. If a time series is too short, we are likely to make
incorrect conclusions about long-term trends. In addition, a time series
that is too long may be a poor use of already limited funds (Gerber,
DeMaster, and Kareiva 1999). Future work can examine other species, with
a wider range of life history characteristics. In addition, similar
approaches can be used to determine the minimum time series length
required to address other questions.

@@

\hypertarget{ref-Johnson2015}{}
Johnson, Paul C.D., Sarah J.E. Barry, Heather M. Ferguson, and Pie
Müller. 2015. ``Power analysis for generalized linear mixed models in
ecology and evolution.'' \emph{Methods in Ecology and Evolution} 6 (2):
133--42.
doi:\href{https://doi.org/10.1111/2041-210X.12306}{10.1111/2041-210X.12306}.

@@

\hypertarget{ref-Myhrvold2015}{}
Myhrvold, Nathan P., Elita Baldridge, Benjamin Chan, Dhileep Sivam,
Daniel L. Freeman, and S.K. Morgan Ernest. 2015. ``An amniote
life-history database to perform comparative analyses with birds,
mammals, and reptiles.'' \emph{Ecology} 96 (11): 3109.
